A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-20 paper.tex
More tidy up

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practise in coalescent theory.
% A ``genome'' in our usage corresponds to the
% haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local trees.
}
\end{figure}

@@

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Figure~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
% are not explicitly concerned with ``coalescence'' (nodes at which
% one or more samples find a common ancestor); for example, node \noderef{e}
% is present but not coalescent in the left-most tree of
% Figure~\ref{fig-arg-in-pedigree}
% (see section XXX for more discussion).
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

% In this section we have considered the nodes in a gARG as separate from events
% in a generative process. Nevertheless, gARGs are commonly constructed to
% include only those genomes associated with specific events. For example,
% \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
% contain[ing] only those edges along which inherited segments of DNA have been
% transmitted, and only those nodes corresponding to ancestors in which there was
% a recombination or coalescence event.''. In an eARG however, events are even
% more explicitly embedded in the data structure.

% There are some interesting and important consequences to this focus on genomes
% rather than events.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
% The difference between eARGs and gARGs are subtle, and whether
% we choose to interpret nodes as events or genomes
% might seem of little practical importance.
% However, there are substantial practical benefits
% from using the gARG encoding, allowing differential
% levels of information about the recombination process to be
% represented in a computationally efficient way.
% The key to this is not so much in how we interpret nodes
% (see Appendix XXX), but specifically in how
% we represent the process of genetic inheritance.
% The critical distinction is that
% the gARG encoding allows us to directly describe
% genetic inheritance in multiple non-contiguous genome intervals
% between an ancestor and a descendant, something that
% cannot be done by storing recombination breakpoints.

@@

\begin{figure}
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks. The vitally important and fundamental distinction between
a gARG and the classical eARG perspective is that gARGs do not
\emph{require} us to posit explicit recombination events in order
to describe a dataset, but
instead provide a means of encoding the effects of
multiple events, and to systematically describe the uncertainty around them
(see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

% JK: previous heading was jargony - this is clearer
\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

>>>>

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practise in coalescent theory.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

\begin{figure}
\begin{center}
    % FIXME making this 0.75 textwidth for now because it's too tall.
    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

@@

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Figure~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

@@

@@

\begin{figure}
% FIXME add local trees in B
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/320
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
Thus, we have the means of encoding the effects of
multiple \emph{simultaneous} events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.
